Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-26 VMCAI-revised-version/final.tex
up

This all changes when one considers VASS (having control-states). For a VASS $V = (Q,T)$, a set $\Bad = \mathop{\uparrow} \Bad$, for all $q \in Q$, either there is an element of the basis of $\Bad$ with control-state $q$ \----
then $q(\textbf{v})$ with $\textbf{v} > \textbf{v}_{\Bad}$ is necessarily in $\Bad$ by upward closure; if this hold for all $q \in Q$, then $k$-resilience does not hold \---- either there is none. If there is none, then 
$\mathop{\uparrow} q(\textbf{0})$
 is in the complement of $\Bad$, i.e. $\Safe$. One can then exhibit upward-closed subsets of 
 $\Safe$ and compute basis for the sets of elements from which they are reachable in at most $k$ steps. 
Substracting these predecessor from $\Bad$ yields either a finite number of elements from which one has to check $\Safe$ is reachable in at most $k$ steps, either an infinite number of elements of which there is one which cannot reach $\Safe$ in at most $k$ steps \-- for much the same reasons {\sc Bounded resilience} and {\sc $k$-resilience} never hold for VAS when $\Safe = \mathop{\downarrow} \Safe$ and $\Bad = \mathop{\uparrow} \Bad$. This lead to a decision procedure
 which lead to the following decidability result:
\begin{theorem}\label{vass down}
{\sc $k$-resilience }  and {\sc Bounded resilience} are decidable for VASS when 
$\Safe = \mathop{\downarrow} \Safe$.
\end{theorem}

\begin{proof}
\iffalse
Let $V$ be a given $d$-VASS and consider a given $k \in \N$.
Consider $\Bad \subseteq Q \times \N^d$ upward-closed and 
$\Safe = S \setminus \Bad$
 downward-closed.
Let us call $c_{\max}$ the maximal absolute value of a constant appearing in a coordinate of a transition.

For all $q \in Q$, either there is an element of the basis of $\Bad$ with control-state $q$ \----
then $q(\textbf{v})$ with $\textbf{v} > \textbf{v}_{\Bad}$ is necessarily in $\Bad$ by upward closure; if this hold for all $q \in Q$, then $k$-resilience does not hold \---- either there is none. If there is none, then 
$\mathop{\uparrow} q(\textbf{0})$
 is in the complement of $\Bad$, i.e. $\Safe$.

From $k_0$ onwards,
$\Bad \setminus (\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))= 
\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is stationnary. 
If the set is finite then we stop the procedure once we have checked for every element
whether it is possible to reach $\Safe$ from it.
If it is possible to reach $\Safe$ from every element then {\sc $k_m$-resilience}
hold
for $k_m = \max(k_0, k_{\pi})$ with 
$k_{\pi}$ the maximum of the length of the paths from 
$\Bad \setminus (\pred^{\leq k_0}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k_0}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$ to $\Safe$.
If it is infinite then
for every $k \in \N$,
$k$-resilience does not hold, 
and we eventually find a witness that $k$-resilience does not hold.
Indeed, 
if
$\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \cup \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is
infinite
then
there exists an element of the form $q(\textbf{u} )$
with $\textbf{u} > \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, \ldots, c_{\max})$ 
that is in $\Bad$ but not in any of the $\pred^{\leq k}(\mathop{\uparrow} q_{j_i}(\textbf{0}))$,
and hence from which it is not possible at all to reach
$\Safe$ in $k$ or less steps. 
\fi
The set $\Bad$ admits a finite basis 
$B_\Bad = \{ q_{i_1}(\textbf{v}_{i_1}), q_{i_2}(\textbf{v}_{i_2}), \ldots,
q_{i_m}(\textbf{v}_{i_m}) \}$.
Consider the vector $\textbf{v}_{\Bad} = \Sigma_{1 \leq j \leq m} \textbf{v}_{i_j}$,
and then consider the vector
$\textbf{u}_k = \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, c_{\max}, \ldots, c_{\max})$.
For all $p \in Q$, all states reachable from $p(\textbf{u}_k)$ in $k$ or less steps are 
of the form $q(\textbf{v})$ with $ \textbf{v} > \textbf{v}_{\Bad}$.

For all $q \in Q$, either there is an element of the basis of $\Bad$ with control-state $q$ \----
then $q(\textbf{v})$ with $\textbf{v} > \textbf{v}_{\Bad}$ is necessarily in $\Bad$ by upward closure; if this hold for all $q \in Q$, then $k$-resilience does not hold \---- either there is none. If there is none, then 
$\mathop{\uparrow} q(\textbf{0})$
 is in the complement of $\Bad$, i.e. $\Safe$.

From then on, we partition $\Safe$ into two subsets: an upward-closed 
union of sets of the form $\mathop{\uparrow} q(\textbf{0})$, and a remaining downward-closed subset.
The key now is that
there is only a finite number of elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} \leq \textbf{u}_k$ for all of which we will check reachability of $\Safe$ in $k$ or less steps, while 
 elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} > \textbf{u}_k$ can only potentially reach the upward-closed subset of $\Safe$ in $k$ or less steps. Only being concerned by the upward-closed subset of $\Safe$ in the latter case enable us to reuse techniques seen when $\Safe$ is upward-closed, in particular the set of
 elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} > \textbf{u}_k$
 and the set of elements from which the upward-closed subset of $\Safe$ is reachable
 are both upward-closed sets a basis of which we can construct, and hence can be checked for inclusion. We refer to appendix~\ref{appendix vass} for more details.\qed
\end{proof}

@@

From $k_0$ onwards,
$\Bad \setminus (\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))= 
\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is stationnary. 
\mathieu{
We check the finiteness of the set by comparing the finite basis of $\Bad$ and
$(\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$.
}
If the set is finite then we stop the procedure once we have checked for every element
whether it is possible to reach $\Safe$ from it.
If it is possible to reach $\Safe$ from every element then {\sc $k_m$-resilience}
hold
for $k_m = \max(k_0, k_{\pi})$ with 
$k_{\pi}$ the maximum of the length of the paths from 
$\Bad \setminus (\pred^{\leq k_0}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k_0}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$ to $\Safe$.
If it is infinite then
for every $k \in \N$,
$k$-resilience does not hold, 
and we eventually find a witness that $k$-resilience does not hold.
Indeed, 
if
$\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \cup \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is
infinite
then
there exists an element of the form $q(\textbf{u} )$
with $\textbf{u} > \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, \ldots, c_{\max})$ 
that is in $\Bad$ but not in any of the $\pred^{\leq k}(\mathop{\uparrow} q_{j_i}(\textbf{0}))$,
and hence from which it is not possible at all to reach
$\Safe$ in $k$ or less steps.
\qed
\end{proof}

>>>>

This all changes when one considers VASS (having control-states). For a VASS $V = (Q,T)$, a set $\Bad = \mathop{\uparrow} \Bad$, for all $q \in Q$, either there is an element of the basis of $\Bad$ with control-state $q$ \----
then $q(\textbf{v})$ with $\textbf{v} > \textbf{v}_{\Bad}$ is necessarily in $\Bad$ by upward closure; if this hold for all $q \in Q$, then $k$-resilience does not hold \---- either there is none. If there is none, then 
$\mathop{\uparrow} q(\textbf{0})$
 is in the complement of $\Bad$, i.e. $\Safe$.
 From then on, we partition $\Safe$ into two subsets: an upward-closed 
union of sets of the form $\mathop{\uparrow} q(\textbf{0})$, and a remaining downward-closed subset.
The key now is that
there is only a finite number of elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} \leq \textbf{u}_k$ for all of which we will check reachability of $\Safe$ in $k$ or less steps, while 
 elements of $\Bad$ of the form $p(\textbf{v})$ with $ \textbf{v} > \textbf{u}_k$ can only potentially reach the upward-closed subset of $\Safe$ in $k$ or less steps. 
Only being concerned by the upward-closed subset of $\Safe$ in the latter case enable us to reuse techniques seen when $\Safe$ is upward-closed, in particular we can compute the basis for the sets of elements from which the upward-closed subset of $\Safe$ is reachable in at most $k$ steps. 
Substracting these predecessor from $\Bad$ yields either a finite number of elements from which one has to check $\Safe$ is reachable in at most $k$ steps, either an infinite number of elements of which there is one which cannot reach $\Safe$ in at most $k$ steps \-- for much the same reasons {\sc Bounded resilience} and {\sc $k$-resilience} never hold for VAS when $\Safe = \mathop{\downarrow} \Safe$ and $\Bad = \mathop{\uparrow} \Bad$. This lead to a decision procedure detailed in appendix~\ref{appendix vass},
 which lead to the following decidability result:
\begin{theorem}\label{vass down}
{\sc $k$-resilience }  and {\sc Bounded resilience} are decidable for VASS when 
$\Safe = \mathop{\downarrow} \Safe$.
\end{theorem}

@@

From $k_0$ onwards,
$\Bad \setminus (\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))= 
\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is stationnary. 
%\mathieu{
We check the finiteness of the set by comparing the finite basis of $\Bad$ and
$(\pred^{\leq k}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$.
%}
If the set is finite then we stop the procedure once we have checked for every element
whether it is possible to reach $\Safe$ from it.
If it is possible to reach $\Safe$ from every element then {\sc $k_m$-resilience}
hold
for $k_m = \max(k_0, k_{\pi})$ with 
$k_{\pi}$ the maximum of the length of the paths from 
$\Bad \setminus (\pred^{\leq k_0}(\mathop{\uparrow} q_{j_1}(\textbf{0}))\cup \ldots \cup \pred^{\leq k_0}(\mathop{\uparrow} q_{j_\ell}(\textbf{0}))$ to $\Safe$.
If it is infinite then
for every $k \in \N$,
$k$-resilience does not hold, 
and we eventually find a witness that $k$-resilience does not hold.
Indeed, 
if
$\Bad \setminus (\pred^{*}(\mathop{\uparrow} q_{j_1}(\textbf{0})) \cup \ldots \cup \pred^{*}(\mathop{\uparrow} q_{j_\ell}(\textbf{0})) )$
is
infinite
then
there exists an element of the form $q(\textbf{u} )$
with $\textbf{u} > \textbf{v}_{\Bad} + (k+1) \cdot (c_{\max}, \ldots, c_{\max})$ 
that is in $\Bad$ but not in any of the $\pred^{\leq k}(\mathop{\uparrow} q_{j_i}(\textbf{0}))$,
and hence from which it is not possible at all to reach
$\Safe$ in $k$ or less steps.
\qed
\end{proof}
